---
title: "Data Tidying"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load in packages

```{r, message = F}
library(dplyr)
library(tidyr)
library(readr)
```

# Read in data

Data come from [Mike Byerly. Alaska commercial salmon catches by management region (1886- 1997). Gulf of Alaska Data Portal. df35b.304.2.](https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1)

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1",
                           show_col_types = F)

head(catch_original)
```

```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)
```


# Clean and reshape data

Drop notes and all column

```{r}
catch_data <- catch_original %>%
  select(-All, -notesRegCode)

head(catch_data)
```

Coerce Chinook column to be numeric instead of character.  OCR failed for row 409.  Replace I with 1

```{r}
catch_clean <- catch_data %>%
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>%
  mutate(Chinook = as.numeric(Chinook))

summary(catch_clean)
```

## Examine NA values

```{r}
i <- which(is.na(catch_clean$Chinook))
i
catch_original[i,]
```

```{r}
catch_long <- catch_clean %>%
  pivot_longer(cols = -c(Region, Year), names_to = "species", values_to = "catch")

head(catch_long)
```

```{r}
catch_wide <- catch_long %>%
  pivot_wider(names_from = "Region", values_from = "catch")

head(catch_wide)
```

```{r}
catch_long <- catch_long %>%
  rename(catch_thousands = catch) %>%
  mutate(catch = catch_thousands * 1000) %>%
  select(-catch_thousands)

head(catch_long)
```

### Split-apply-combine

```{r}
mean_region <- catch_long %>%
  group_by(Region) %>%
  summarize(catch_mean = mean(catch),
            n = n())

mean_region
```

```{r}
mean_species <- catch_long %>%
  group_by(species) %>%
  summarize(species_mean = mean(catch),
            n = n())

mean_species
```

```{r}
mean_year <- catch_long %>%
  group_by(Year) %>%
  summarize(year_mean = mean(catch),
            n = n())

mean_year
```

```{r}
var <- catch_long %>%
  filter(Year > 1899) %>%
  group_by(species, Year) %>%
  summarize(mean_catch = mean(catch), .groups = "drop") %>%
  arrange(-mean_catch)

head(var)
```

# Join to regions dataset

```{r}
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))

head(catch_joined)
```
...but you can also just rename "code" to "Region" in region_defs